---
title: 'Coursera: Statistical Inference - Course Project'
author: "Konstantin Greger"
date: "2015/01/24"
output:
  html_document: default
  pdf_document: default
---

## Part 2: Analysis of the "ToothGrowth"" Data in the R "datasets"" Package

```{r}
# load the data
library(datasets)
tooth = ToothGrowth
```

The "ToothGrowth" data in the R "datasets" package consists of `r nrow(tooth)` observations. According to the [official description of the dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html) it provides "the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid)."

Tooth length varies between `r min(tooth$len[tooth$supp == "OJ"])` and `r max(tooth$len[tooth$supp == "OJ"])` mm for the celivery using orange juice, and between `r min(tooth$len[tooth$supp == "VC"])` and `r max(tooth$len[tooth$supp == "VC"])` mm for the delivery using ascorbic acid.

The following plot shows the data distribution over delivery methods and supplement doses:

```{r}
# plot the distributions
library(ggplot2)
ggplot(data = tooth, 
       aes(x = as.factor(dose), 
           y = len, 
           fill = supp)) +
    geom_bar(stat = "identity") +
    facet_grid(. ~ supp) +
    xlab("Supplement dose in mg") +
    ylab("Tooth length in mm") +
    guides(fill = guide_legend(title = "Supplement type"))
```

The data show a clear positive correlation between tooth length and the dose levels of Vitamin C for both delivery methods. The supplement dose appears to be of greater importance when the supplement is delivered as ascorbic acid as compared to orange juice.

The null hypothesis $H_0$ is that the supplement type of vitamin C generally does not have an effect on the tooth length.

```{r}
# check for differences due to different supplement type
# assumption: unequal variances between the two groups
t.test(len ~ supp, 
       data = tooth)
```

With a p-value of `r round(t.test(len ~ supp, data = tooth)$p.value, 2)` and a confidence interval containing zero we can not reject the null hypothesis $H_0$ that the different supplement types have no effect on tooth length.

It is worthwhile to perform pairwise comparisons between the different supplement doses:

```{r}
# create subsets for the supplement dose level pairs
tooth_0.5_1.0 <- subset(tooth, dose %in% c(0.5, 1.0)) 
tooth_0.5_2.0 <- subset(tooth, dose %in% c(0.5, 2.0)) 
tooth_1.0_2.0 <- subset(tooth, dose %in% c(1.0, 2.0)) 
```

```{r}
# check for differences due to different dose levels: 0.5, 1.0
# assumption: unequal variances between the two groups
t.test(len ~ dose, 
       data = tooth_0.5_1.0)
```

```{r}
# check for differences due to different dose levels: 0.5, 2.0
# assumption: unequal variances between the two groups
t.test(len ~ dose, 
       data = tooth_0.5_2.0)
```

```{r}
# check for differences due to different dose levels: 1.0, 2.0
# assumption: unequal variances between the two groups
t.test(len ~ dose, 
       data = tooth_1.0_2.0)
```

For all three dose level pairs (0.5 and 1.0 mg, 0.5 and 2.0 mg, 1.0 and 2.0 mg), the p-values are smaller than 0.05 (`r round(t.test(len ~ supp, data = tooth_0.5_1.0)$p.value, 2)`, `r round(t.test(len ~ supp, data = tooth_0.5_2.0)$p.value, 2)`, and `r round(t.test(len ~ supp, data = tooth_1.0_2.0)$p.value, 2)`, respectively), and none of the confidence intervals does contain zero. Therefire we can reject the null hypothesis $H_0$ and establish that increasing the dose level leads to an increase in tooth length.

All analyses were performed under the assumption of a random assignment of guinea pigs to different dose level categories and supplement types, and also the representativeness of the samples.